ODGI: understanding pangenome graphs

### 2021-10-27 manuscript/fig_metrics.tex
one column-width metrics figure

\begin{figure*}[h!]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth, trim=0 1cm 0 0.5cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
		\label{fig:metrics-multiqc}
	\end{subfigure}
	\begin{subfigure}{1\linewidth}
	\caption{}
	\centering
	% include fourth image
	\includegraphics[width=\linewidth,trim=+.225cm 2.2cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
	\label{fig:metrics-degree}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
	\caption{}
	\centering
	% include second image
	\includegraphics[width=\linewidth,trim=+.225cm 1.2cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
	\label{fig:metrics-depth}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth, trim=-1.5cm 0.75cm -0.8cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
		\label{fig:metrics-viz}
	\end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Features of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module. The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}). \textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph. Around position 200 there is a huge variation in node degree. \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph. The alternating depth around position 200 indicates polymorphic variation complementing the above node degree analysis. \textbf{(d)} \textit{odgi viz} visualization of the 13 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph. Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
	\label{fig:metrics}
\end{figure*}

>>>>

\begin{figure}[h!]
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include first image
		\includegraphics[width=1.0\linewidth, trim=-1.25cm 3cm 0 1.75cm]{fig/metrics/chr4_pan_HTTex1_gfa_multiqc_odgi_stats}
		\label{fig:metrics-multiqc}
	\end{subfigure}
	\begin{subfigure}{1\linewidth}
	\caption{}
	\centering
	% include fourth image
	\includegraphics[width=\linewidth,trim=+.225cm 3cm +.425cm +3cm]{fig/metrics/chr4_HTT_chm13_degree_w1_bed}
	\label{fig:metrics-degree}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
	\caption{}
	\centering
	% include second image
	\includegraphics[width=\linewidth,trim=+.225cm 3cm +0.425cm +3cm]{fig/metrics/chr4_HTT_chm13_depth_w1_bed}
	\label{fig:metrics-depth}
	\end{subfigure}
	\begin{subfigure}{\linewidth}
		\caption{}
		\centering
		% include second image
		\includegraphics[width=1.0\linewidth, trim=-1.5cm 0.75cm -0.8cm 0.5cm]{fig/metrics/chr4_pan_fa_a2fb268_e820cd3_9ea71d8_smooth_gfa_og_HTTex1_og_O_og_tiny_og_png_svg}
		\label{fig:metrics-viz}
	\end{subfigure}
%	\includegraphics[width=\linewidth]{fig/metrics/chr4.pan.HTTex1.gfa.multiqc_odgi_stats.png}
	\caption{Features of a 90 haplotypes human pangenome graph of the exon 1 huntingtin gene (\textit{HTTexon1}): \textbf{(a)} Excerpt of vital statistics of the \textit{HTTexon1} graph displayed by MultiQC's ODGI module. The very high GC content of 73.0\% compared to a human genomic mean GC content of 40.9\% \cite{Piovesan2019} is in accordance with the literature (see for example \cite{Sathasivam2013, Neueder2017}). \textbf{(b)} Per nucleotide node degree distribution of CHM13 in the \textit{HTTexon1} graph. Around position 200 there is a huge variation in node degree. \textbf{(c)} Per nucleotide node depth distribution of CHM13 in the \textit{HTTexon1} graph. The alternating depth around position 200 indicates polymorphic variation complementing the above node degree analysis. \textbf{(d)} \textit{odgi viz} visualization of the 23 largest gene alleles, CHM13, and GRCH38 of the \textit{HTTexon1} graph. Figures \textbf{(b)-(d)} clearly highlight the variant region around position 200 showing the variable number of glutamine residues of the different individuals as reported by \cite{Nance1999}.}
	\label{fig:metrics}
\end{figure}

### 2021-10-27 manuscript/main.tex
forgot citation
one column-width metrics figure

Each tool focuses on a small set of related operations.
Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
This supports the implementation of flexible, composable, graph processing pipelines based on a common data type.
While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
By considering all paths in the graph as potential reference or query sequences, we make the graph invisible to downstream tools that operate on collections of sequences (e.g. bedtools, samtools, aligners).
This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

@@

\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1 format. It can reveal a graph’s internal structures for e.g. debugging processes.

\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph.

@@

ODGI's visualization tools produce static, large-scale snapshots. Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021} or focusing on one dimension \citep{Wick_2015}. However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing. ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.

>>>>

Each tool focuses on a small set of related operations.
Most read or write the native ODGI format (`og' extension) (Figure \ref{fig:operations}) and work with standard text based data formats common to bioinformatics.
This supports the implementation of flexible, composable, graph processing pipelines based on a common data type.
While the pangenome graph is a first-class entity in ODGI, we focus on the embedded paths to provide a universal coordinate system.
By considering all paths in the graph as potential reference or query sequences, we make the graph invisible to downstream tools that operate on collections of sequences (e.g. BEDTools, SAMtools \citep{Li2009}, aligners).
This approach benefits from the information in the graph without strongly embedding our methods in this difficult new research context.

@@

\textit{odgi view} can convert a graph in ODGI binary format to the standard GFAv1\footnote{\url{https://github.com/GFA-spec/GFA-spec/blob/master/GFA1.md} (accessed Oct 2021)}  format. It can reveal a graph’s internal structures for e.g. debugging processes.

\textit{odgi stats} outputs pangenome statistics in TSV or YAML textual file formats.
Among other metrics, it can calculate the number of nodes, edges, paths and the total nucleotide length of the graph. {\color{red} FIXME: All this information is also listed in the metrics section. How to split? No mentioning of file formats in the following sections?}

@@

ODGI's visualization tools produce static, large-scale snapshots. Recent interactive browsers are reference centric \citep{Beyer2019, Yokoyama2019, Durant2021, Liang2021} or focusing on one dimension \citep{Wick_2015, Gonnella2018}. However, an interactive visualization solution that combines the 1D and 2D layout of a graph with annotation and read mapping information across different zoom levels is still missing. ODGI's graph layout algorithms could become a backbone of such tools, as is already the case for \textit{gfaestus}\footnote{\url{https://github.com/chfi/gfaestus}}, which currently relies on the \textit{odgi layout} output for its interactive 2D visualization.
